---
title: "Session 1"
author: "Ottar N. Bjornstad"
date: "2023-11-08"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

RStudio [https://www.rstudio.com] is an interface to the R language which is together with Python is the most widely used programming platform in many fields of computing. It is known for it's extremely broad statistical capabilities: cran.r-project.org serves out more than 10k add-on packages in addition to the R base... However in this workshop we will use it mainly for its raw numerical and computational power. To prep for the workshop, after installing Rstudio we need to install the **epimdr2** and **tinytex** packages either via the menu: ``Tools -> Install packages'' or the console:

```{r eval=FALSE}
install.packages("epimdr2")
install.packages("tinytex")
```

Once a package has been installed, it is installed forever but needs to be attached to be used:
```{r eval=FALSE}
install.packages("epimdr2")
```

There are an enormous amount of online and printed resources for R. One usefull place is a search engine sole devoted to sifting through R related materials [https://rseek.org].

**Six Things**
R may seem complicated, but the language itself is really only made up of six'ish things that is used to build objects.

**Vectors** are made forexample using the c() concatenate function... function are executed using round brackets:
```{r}
x=c(1,5,3, 2, 23, 0)
x
y=rep(1.25, 20)
y
```
The [] indicate the element number for subsetting:
```{r}
x[4]
x[c(2,5)]
```

**Matrices** are 2D arrays (there are higher D arrays also but we wont use them) which we subset by row column. or element:

```{r}
z=matrix(x, nrow=3, byrow=TRUE)
z
z[2,]
z[,1]
z[3,2]
```

**Data.frames** are like simple spread-sheets. Easiest to generate by exporting a sheet to a comma separated  file and then read in (note that names has to be in quotes otherwise R will try to evaluate it as an object), data.frames are accessed using &#36;name or [,2]:
```{r}
meas=read.csv("meas.csv", header=TRUE)
head(meas)
head(meas)$London
head(meas)[,2]
```

**Lists** are unstructured collection objects that can be accessed using &#36;name or [[]] element number:

```{r}
lst=list(a=x, b=y, pip=z)
lst$pip
lst[[1]]
```


**Functions** does things (when followed by () otherwise the function will just be displayed). We can for example use the plot() funtion to plot the measles data:

```{r pressure, echo=FALSE}
plot(London~time, data=meas)
```

The ~ signifies a formula, i.e London as a function of time. We would get the same result from plot(x=meas$&#36;time, y=meas&#36;London). Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

We can also write our own functions which will be very important during the workshop:
```{r}
myfn=function(arg){
ss=sum(arg)
return(ss)
}

myfn(y)
```

Finally, loops are when we want to repeat something many times (we will visit more on this tomorrow):

```{r}
for(i in 1:4){
  cat(x[1:i], "\n")
}
```
The cat() function simply prints to screen and the colon operator generates a vector of a:b whole numbers from a to b.

