---
title: "Session5"
author: "Ottar N. Bjornstad"
date: "2023-11-08"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

COVID-19 has loss of immunity.

The SIRS model is:

\begin{align}
    \frac{dS}{dt} =& \underbrace{\mu N}_{\mbox{birth}} - \underbrace{\beta I \frac{S}{N}}_{\mbox{infection}}  + \underbrace{\omega R}_{\mbox{resusceptible}}- \underbrace{\mu S}_{\mbox{death}} \label{eq:sirss}\\
     \frac{dI}{dt} =& \underbrace{\beta I \frac{S}{N}}_{\mbox{infection}} - \underbrace{\gamma I}_{\mbox{recovery}} - \underbrace{\mu I}_{\mbox{death}}  \label{eq:sirsi}\\
     \frac{dR}{dt} =& \underbrace{\gamma I}_{\mbox{recovery}}  - \underbrace{\omega R}_{\mbox{lost immmunity}} - \underbrace{\mu R}_{\mbox{death}}, \label{eq:sirsr}
\end{align}

where $\omega$ is the rate of loss of immunity ($\sim 0.5 \mbox{ year}^{-1}$),  an infectious period ($1/\gamma$) of 3.8 days and maybe suggets an early $R_0$ of $2.9$. Modeling fractions ($N=1$) the weekly SIRS appropriate parameters are:

```{r}
N = 1
gamma = 7 / 3.8 
omega = 1 / (52 * 4)
mu = 1 / (52 * 70)
R0 = 2.9 
```

The call for back-calculating $\beta$ to get the right $R_0$ and gathering parameters is:

```{r}
#R0 = beta / (gamma + mu)
beta = R0 * (gamma + mu)
paras = c(beta = beta, gamma = gamma, 
   mu = mu, omega = omega)
```


STEP 1:

STEP 2:

STEP 3:

STEP 4:

STEP 5:


Work in groups to see if you can start work on the problem...